---
title: "Tidying Data in R"
author: "Bas Machielsen"
date: \today
institute: Utrecht University
fontsize: 10pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
output:
    beamer_presentation: 
      keep_tex: yes
      theme: Madrid
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
if (!require(tidyverse)) {install.packages('tidyverse', ask = FALSE, checkBuilt = TRUE, type = "binary")}
library(tidyverse)

if (!require(stargazer)) {install.packages('stargazer', ask = FALSE, checkBuilt = TRUE, type = "binary")}
library(stargazer)

if (!require(kableExtra)) {install.packages('kableExtra', ask = FALSE, checkBuilt = TRUE, type = "binary")}
library(kableExtra)

if (!require(RCurl)) {install.packages('RCurl', ask = FALSE, checkBuilt = TRUE, type = "binary")}
library(RCurl)

if (!require(xtable)) {install.packages('xtable', ask = FALSE, checkBuilt = TRUE, type = "binary")}
library(xtable)
options(xtable.comment = FALSE)

```

## What is tidy data?

- In order for your data to be ready for analysis, it needs to be **tidy**. 

- Tidy data is defined as: 

> data sets that are arranged such that each variable is a column and each observation (or case) is a row.

- I will attempt to explain how to put this principle in practice using a set of packages called the **tidyverse**, but strictly speaking, only the `tidyr` package is required. 

- You can download the .Rmd and .csv files from my [**github page**](www.github.com/basm92), and follow along! 

## What is (un)tidy data?

> All tidy data are similar to each other. All untidy data are untidy in their own way.

- Although tidy data shares the properties mentioned in the previous slide, untidy data refers (broadly speaking) to unstructured data. 

- An example (from [www.tidyverse.org](www.tidyverse.org)):

```{r}
preg <- read.csv("https://raw.githubusercontent.com/tidyverse/tidyr/master/vignettes/preg.csv")

kable(preg, caption = "This is what an untidy dataset looks like", 
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```


## An example of untidy data

- Is each obervation in a row?

- Is each variable in a column?

- What are the variables here? 

```{r}
preg2 <- read.csv("https://raw.githubusercontent.com/tidyverse/tidyr/master/vignettes/preg2.csv")

kable(preg2, caption = "This is also what an untidy dataset looks like", 
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```

## Example: from untidy to tidy data

- The variables are (i) the treatment, (ii) the subject, and (iii) the 'score' corresponding to each subject-treatment observations. 

```{r, echo = TRUE, include = TRUE, eval = FALSE}
preg %>%
    pivot_longer(treatmenta:treatmentb, 
                 names_to = "treatment", values_to = "score") %>% 
    mutate(treatment = gsub("treatment", "", treatment)) %>%
    arrange(name, treatment) %>%
    kable(caption = "Test", booktabs = TRUE, 
          row.names = FALSE) %>%
    kable_styling(latex_options = "striped")
```

## Example: from untidy to tidy data

- The former chunk of code transforms the data in the first table to the following table: 

```{r}
preg %>%
    pivot_longer(treatmenta:treatmentb, 
                 names_to = "treatment", values_to = "score") %>% 
    mutate(treatment = gsub("treatment", "", treatment)) %>%
    arrange(name, treatment) %>%
    kable(caption = "This is tidy data", 
          booktabs = TRUE, 
          row.names = FALSE) %>%
    kable_styling(latex_options = "striped")
```


## Transforming the data

- The key command in the former chunk of code is `pivot_longer`. With the arguements data, columns, names_to and values_to. 

- All the other functions are essentially layout changes, not fundamental transformations. 

- The command pivot_longer basically transforms the dataset from: 

```{r}
kable(preg, caption = "This is what an untidy dataset looks like", 
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```

## Transforming the data

- to : 

```{r, echo = TRUE}
pivot_longer(data = preg, cols = 2:3, names_to = "treatment", 
             values_to = "score") %>%
    kable(caption = "This is what a tidy dataset looks like", 
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```

## Contents 

- The best way to see how tidy data works is by using examples. In this lecture, I will demonstrate extensively how to create tidy data that is ready for analysis using two examples.

- First, I will show how to combine and tidy data on economic development from the World Bank.

- Afterwards, I show how to combine and tidy data from Amadeus, a commercial database. 

- On the fly, I will demonstrate how simple it is to analyze or create graphs with tidy data. 

- Finally, I will use other examples to show how to tidy more messy data, inspired by examples from the [**tidyverse website**](https://tidyr.tidyverse.org/reference/pivot_longer.html)

## World Bank Dataset

- Let us now download some data from the World Bank database. I extract the 55 basic World Development Indicators from [**here**](https://databank.worldbank.org/source/world-development-indicators/) for 20 countries. 

- Importing the data.. 

```{r, echo = TRUE, message = FALSE, warning = FALSE}
worldbank <- read_csv("wb.csv")

dim(worldbank)
```

- The data has 1105 observations and 14 variables. That is not at all what we wanted. What do the data look like?

## World Bank Dataset

```{r, echo = TRUE, message = FALSE, warning = FALSE}
kable(worldbank[1:4,c(1,2,4,5)], 
      caption = "Work Bank Data - Untidy!", 
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```

- Very untidy dataset! NA observations are entered as `..`, and variable names require an extensive definition before you know what they mean. 

- Furthermore, years are not notated straightforwardly, and the data violates the tidy data principles. 

## What to do?

- First, let us try to save the variable names and series code to another dataset (step 1), and delete the names from the worldbank dataset (step 2). 

```{r, echo = TRUE}
#Step 1
wbnames <- cbind(unique(worldbank[,3]),unique(worldbank[,4])) %>%
    na.omit()

#Step 2
worldbank <- worldbank[1:1100,-3]
```

## Almost there..

- Now, let's try to unpivot the columns containing the years.. 


```{r, echo = TRUE, results = FALSE}
worldbank <- pivot_longer(worldbank, 4:13, 
             names_to = "years", 
             values_to = "value")
```

- This is what the data looks like right now. 

```{r}
kable(worldbank[1:4,2:5], 
      caption = "The data then looks like this.", 
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```


## Final steps

- Let's now clean the "years" variable (step 1), set the `..` observations to NA (step 2), and "widen" the data so that every separate indicator gets a column (step 3). 

- Step 3 is done using the `pivot_wider` command: it widens the data and shortens the number of observations, because it transfers information from rows to columns. 

```{r, echo = TRUE}
#Step 1
worldbank <- worldbank %>%
    mutate(years = substring(years, 1, 4))

#Step 2
worldbank$value[worldbank$value == ".."] <- NA

#Step 3
worldbank <- pivot_wider(data = worldbank,
                         names_from = `Series Code`,
                         values_from = `value`)
```


## Done! 

- `pivot_wider`, takes three arguments: the data, from which column it should take the new column names (names_from), and from which column it should take the values belonging to the corresonding cels (values_from). 

- This is what it looks like! 

```{r}
kable(worldbank[1:4,1:5], caption = "Tidy Data",
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")

```

## Military Spending

- Suppose I want to know the military expenditures as a percentage of GDP `r wbnames[30,2]` for each country in the dataset in 2015. 

- I can use the `wbnames` information we've saved to add labels to the data. 

```{r, echo = TRUE, include = TRUE, eval = FALSE}
worldbank %>%
    filter(years == "2015") %>%
    ggplot(aes(x = `Country Code`, y = `MS.MIL.XPND.GD.ZS`)) + 
    geom_col() + 
    labs(y = as.character(wbnames[match("MS.MIL.XPND.GD.ZS",
                                        wbnames[,2]),1])) +
    theme(axis.text.y =element_blank())

```

## Military Spending

```{r, echo = FALSE, fig.height = 2.5, fig.width = 6}
worldbank %>%
    filter(years == "2015") %>%
    ggplot(aes(x = `Country Code`, y = `MS.MIL.XPND.GD.ZS`)) + 
    geom_col() + 
    labs(y = as.character(wbnames[match("MS.MIL.XPND.GD.ZS",
                                        wbnames[,2]),1])) +
    theme(axis.text.y =element_blank())

```

## Fertility Rate

- Suppose now you  want to track the evolution of the fertility rate over the last 10 years of some countries in the dataset. 

```{r, echo = TRUE, include = TRUE, eval = FALSE, warning = FALSE}
worldbank %>%
    filter(years < 2018, 
           `Country Code` == "CHE" | 
             `Country Code` == "GBR"
           | `Country Code` == "RUS" | 
             `Country Code` == "SAU") %>%
    ggplot(aes(x = years, y = `SP.DYN.TFRT.IN`, 
               group = `Country Code`, 
               color = `Country Code`)) + 
    geom_line() + 
    labs(y = as.character(wbnames[match("SP.DYN.TFRT.IN",
                                        wbnames[,2]),1])) +
    theme(axis.text.y =element_blank(), 
          axis.ticks.y = element_blank())
```

## Fertility Rate

```{r, echo = FALSE, warning = FALSE, fig.height = 2.5, fig.width = 6}
worldbank %>%
    filter(years < 2018, 
           `Country Code` == "CHE" | 
             `Country Code` == "GBR"
           | `Country Code` == "RUS" | 
             `Country Code` == "SAU") %>%
    ggplot(aes(x = years, y = `SP.DYN.TFRT.IN`, 
               group = `Country Code`, 
               color = `Country Code`)) + 
    geom_line() + 
    labs(y = as.character(wbnames[match("SP.DYN.TFRT.IN",
                                        wbnames[,2]),1])) +
    theme(axis.text.y =element_blank(), 
          axis.ticks.y = element_blank())
```

## Another untidy dataset..

- Now, I will download some data from Amadeus, a commercial database collected information about firms across the world. 

- I collect data from 2000 European companies on their financial information, the number of employees, and the number of directors/managers. 

```{r, echo = TRUE}
amadeus <- read.csv("Amadeus_Export_1.csv", fileEncoding = "UCS-2LE")

dim(amadeus)
```

## What does the data look like?

- These are some of the variables.. 

```{r}
kable(names(amadeus[1:9]), caption = "Variable names",
      booktabs = TRUE, row.names = FALSE, col.names = "") %>%
  kable_styling(latex_options = "striped")
```

## Tidying the Amadeus data

- The data look very untidy. Column names contain variables and years, and there are various superfluous columns, such as `Mark`, and last year. NA observations are coded as `n.a.` but are not recognized by are as such. 

- The first thing we need to do is to change the NA obs to functioning NA obs, and remove some superfluous columns

```{r, echo = TRUE}
#Step 1
amadeus[amadeus == "n.a."] <- NA
amadeus[amadeus == "n.s."] <- NA
amadeus <- amadeus[,-c(1,5:7)]
```

## Using pivot to transform the data

- Next, we want to convert the data into a __long__ format, using `pivot_longer`. 

```{r, echo = TRUE}
amadeus <- amadeus %>%
  mutate_all(as.character)

amadeus <- pivot_longer(data = amadeus,
                        cols = 4:64,
                        names_to = "variable",
                        values_to = "value")

```

## Using pivot to transform the data

- This is what the data looks like now: 

```{r, echo = FALSE, warning = FALSE}
kable(names(amadeus[1:5]), caption = "The amadeus dataset",
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```

## String correction

- Now, we correct the strings using the following steps: 

```{r, echo = TRUE, warning = FALSE}
#First, remove "th.EUR" from the string
amadeus$variable <- sub("\\.th.EUR.","", 
                        amadeus$variable)
#Extract the year and put it into a new variable
amadeus$year <- as.numeric(
  str_extract(amadeus$variable, "[0-9]+"))

#Include only letters from the alphabet as variable names
amadeus$variable <- sapply(
  str_extract_all(
      amadeus$variable,"[aA-zZ]+"), 
  paste, collapse = "_")
```


## Pivoting...

- Finally, we can expand the table to adhere to the principal of one obs. per row, one variable per column.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
amadeus <- pivot_wider(data = amadeus, names_from = variable, 
                       values_from = value)

#There is only one problem, No. of directors/managers 
#violates the tidy data principles.
nodir <- amadeus %>%
  filter(is.na(year))
amadeus <- amadeus %>%
  filter(!is.na(year))

amadeus <- merge(amadeus[,1:16], 
                 nodir[,c(1,17)], 
                 by = 1)

amadeus[,5:17] <- data.frame(sapply(amadeus[,5:17], 
                                    function(x) as.numeric(x)))
```

## Done!

- Now, the data looks like this:

```{r}
kable(amadeus[1:5,c(1,4:5)], caption = "Tidy data",
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "striped")
```

- All other variables are located in the remaining columns.

## Some analyses

## Summarizing

- It seems that cleaning this type of data already required a fair amount of work, including some programming that requires knowledge of regular expressions, and some fairly non-standard transformation.

- In the remainder, I show some other examples of `pivot_longer` and `pivot_wider` in non-standard situations, so you do not get stuck when coming across these types of data. 





